\documentclass[a4paper]{article}
\usepackage[margin=1in]{geometry}

\usepackage{preamble}
\usepackage{log221-a4-macros}
\usepackage[final]{showkeys}

\title{LOG 221 - A4}
\author{Frank Tsai}

\begin{document}

\maketitle

\begin{lemma}\label{0000}
  If $\dVdash{n}{\ell}{k}{q} \Gamma \Rightarrow C,\Delta$ and $\dVdash{m}{\ell}{k}{q} \Lambda,C \Rightarrow \Pi$ with $\dRank{C} = \ell + 1$, $\cRank{C} < k$, and $\qRank{C} < q$, then $\dVdash{h}{\ell}{k}{q} \Gamma,\Lambda \Rightarrow \Delta,\Pi$, where $h \leq \max(n,m) + 2$.
\end{lemma}
\begin{proof}
  We proceed by case analysis on $C$.
  Since $\dRank{C} \geq 1$, we do not have to consider atoms, $\bot$, and $Qx.A(x)$.
  \begin{itemize}
  \item[$\lnot A$:] By inversion, we have $\dVdash{n}{\ell}{k}{q} \Gamma,A \Rightarrow \Delta$ and $\dVdash{m}{\ell}{k}{q} \Gamma \Rightarrow A, \Delta$.
    Since $\dRank{A} = \ell$, we can afford using \rCut{} to obtain the required sequent.
    The height of this derivation is bounded by $\max(n,m) + 1$.
  \item[$A \wedge B$:] By inversion, we have $\dVdash{n}{\ell}{k}{q} \Gamma \Rightarrow A,\Delta$; $\dVdash{n}{\ell}{k}{q} \Gamma \Rightarrow B,\Delta$; and $\dVdash{m}{\ell}{k}{q} \Lambda,A,B \Rightarrow \Pi$.
    Since $\dRank{A}$ and $\dRank{B}$ are bounded by $\ell$, we can afford two \rCut{}s to obtained the required sequent.
    \[
      \ebrule{
        \hypo{\Gamma \Rightarrow A,\Delta}
        \hypo{\Gamma \Rightarrow B,\Delta}
        \hypo{\Lambda,A,B \Rightarrow \Pi}
        \infer2[\rCut]{\Gamma,\Lambda,A \Rightarrow \Delta,\Pi}
        \infer2[\rCut]{\Gamma,\Lambda \Rightarrow \Delta,\Pi}
      }
    \]
    The height of this derivation is bounded by $\max(n,m) + 2$.
  \item[$A \vee B$:] This is analogous to the previous case: apply inversion on both assumptions; and then apply two \rCut{}s.
  \end{itemize}
\end{proof}

\section*{Problem 1}\label{0001}
\begin{proof}[Solution]
  Observe that the proof of \cref{0000} goes through even if we weaken the assumption to allow $\cRank{C} = k$ since inversion yields strictly smaller formulas.
\end{proof}

\section*{Problem 2}\label{0009}

\begin{lemma}\label{0004}
  If $\dVdash{n}{0}{k}{q} \Gamma \Rightarrow C,\Delta$ and $\dVdash{m}{0}{k}{q} \Lambda,C \Rightarrow \Pi$ with $\cRank{C} < k$ and $\qRank{C} = q$, then $\dVdash{h}{k-1}{k-1}{q} \Gamma,\Lambda \Rightarrow \Delta,\Pi$, where $h \leq n + m$.
\end{lemma}
\begin{proof}
  We proceed by induction on $n + m$; there are two cases to consider: (1) $C$ is principal in both assumptions, or (2) $C$ is not principal in at least one assumption.

  We treat the first case by case analysis on $C$.
  Note that $\bot$ does not apply in this case since it does not have a right rule; hence there are only three cases to consider:
  \begin{itemize}
  \item[Atom:] In this case we have $C \in \Gamma$ and $C \in \Pi$; hence $\dVdash{0}{0}{0}{0} \Gamma,\Lambda \Rightarrow \Delta,\Pi$ by \rId.
  \item[$\forall x.A(x)$:] By inversion, we have $\dVdash{n}{0}{k}{q} \Gamma \Rightarrow A(s),\Delta$ for any $s$.
    Unlike \cref{0000}, the lack of left inversion means we must handle implicit contraction correctly.

    We apply the induction hypothesis to $\dVdash{n}{0}{k}{q} \Gamma \Rightarrow \forall x.A(x), \Delta$ and $\dVdash{m-1}{0}{k}{q} \Lambda,A(t),\forall x.A(x) \Rightarrow \Pi$ to eliminate $\forall x.A(x)$.
    This yields $\dVdash{n+m-1}{k-1}{k-1}{q} \Gamma,\Lambda,A(t) \Rightarrow \Delta,\Pi$.

    Since $\qRank{A(t)} < \qRank{\forall x.A(x)} < q$ and $\cRank{A(t)} < \cRank{\forall x.A(t)} < k$, we can afford to use \rCut{} to remove $A(t)$:
    \[
      \ebrule{
        \hypo{\Gamma,\Lambda,A(t) \Rightarrow \Delta,\Pi}
        \hypo{\Gamma \Rightarrow A(t),\Delta}
        \infer2[\rCut]{\Gamma,\Lambda \Rightarrow \Delta,\Pi}
      }
    \]
    Note that $d$-rank is bounded by the cut rank; hence the resulting derivation satisfies all complexity bounds.
  \item[$\exists x.A(x)$:] Completely analogous to the previous case.
    This time, we lack right inversion, so we handle implicit contraction by applying the induction hypothesis.
  \end{itemize}
  In the latter case, we may assume, by symmetry of the argument, that $C$ is not principal in the first assumption.
  Every sequent calculus rule has the following general form:
  \begin{mathpar}
    \ebrule[*]{
      \hypo{\Gamma_1 \Rightarrow \Delta_1}
      \hypo{\Gamma_2 \Rightarrow \Delta_2}
      \infer2{\Gamma \Rightarrow \Delta}
    }
  \end{mathpar}
  Hence we have $\dVdash{n-1}{0}{k}{q} \Gamma_1 \Rightarrow C,\Delta_1$ and $\dVdash{n-1}{0}{k}{q} \Gamma_2 \Rightarrow C,\Delta_2$.
  Then by the induction hypothesis, we have $\dVdash{n+m-1}{k-1}{k-1}{q} \Gamma,\Lambda_1 \Rightarrow \Delta,\Pi_1$ and $\dVdash{n+m-1}{k-1}{k-1}{q} \Gamma,\Lambda_2 \Rightarrow \Delta,\Pi_2$.
  Now apply *.
\end{proof}

\begin{lemma}\label{0002}
  If $\dVdash{n}{\ell + 1}{k}{q} \Gamma \Rightarrow \Delta$, then $\dVdash{2n}{\ell}{k}{q} \Gamma \Rightarrow \Delta$.
\end{lemma}
\begin{proof}
  We proceed by induction on $n$; and we do a case analysis on the last applied rule.
  Most cases are uninteresting: apply the induction hypothesis to the premise(es) and then the corresponding rule.
  We present a few representative cases for completeness.
  The only interesting case is \rCut.
  \begin{itemize}
  \item[\rId:] In this case, there is no cut formula; hence we already have $\dVdash{2n}{\ell}{k}{q} \Gamma \Rightarrow \Delta$.
  \item[\rLBot:] Analogous to \rId.
  \item[\rRNot:] The induction hypothesis yields $\dVdash{2n - 2}{\ell}{k}{q} \Gamma, A \Rightarrow \Delta$.
    Then we obtain $\dVdash{2n - 1}{\ell}{k}{q} \Gamma \Rightarrow \lnot A, \Delta$ by applying \rRNot.
    This is clearly bounded by $2n$.
  \item[\rLDis:] The induction hypothesis yields $\dVdash{2n - 2}{\ell}{k}{q} \Gamma, A \Rightarrow \Delta$ and $\dVdash{2n - 2}{\ell}{k}{q} \Gamma, B \Rightarrow \Delta$, then we obtain $\dVdash{2n - 1}{\ell}{k}{q} \Gamma, A \vee B \Rightarrow \Delta$ by applying \rLDis.
  \item[\rCut:] Let $C$ be the cut formula.
    By the induction hypothesis, we have $\dVdash{2(n-1)}{\ell}{k}{q} \Gamma \Rightarrow C,\Delta$ and $\dVdash{2(n-1)}{\ell}{k}{q} \Gamma,C \Rightarrow \Delta$.
    There are two cases to consider: (1) $\dRank{C} < \ell + 1$, or (2) $\dRank{C} = \ell + 1$.
    In the former case, we do not have to alter anything since we can afford this \rCut, while in the latter case, we apply \cref{0000}, yielding $\dVdash{2n}{\ell}{k}{q} \Gamma \Rightarrow \Delta$.
  \end{itemize}
\end{proof}

By applying \cref{0002} recursively on $\ell$, we get \cref{0003}.

\begin{corollary}\label{0003}
  If $\dVdash{n}{\ell}{k}{q} \Gamma \Rightarrow \Delta$, then $\dVdash{2^{\ell}n}{0}{k}{q} \Gamma \Rightarrow \Delta$.
\end{corollary}

\begin{lemma}\label{0005}
  If $\dVdash{n}{0}{k}{q+1} \Gamma \Rightarrow \Delta$, then $\dVdash{2^{n \cdot k}}{k-1}{k-1}{q} \Gamma \Rightarrow \Delta$.
\end{lemma}
\begin{proof}
  Similar to \cref{0002}, most cases are uninteresting.
  We look at \rCut.
  Let $C$ be the cut formula.
  There are two cases to consider: (1) $\qRank{C} < q$, or (2) $\qRank{C} = q$.
  In the former case, nothing needs to be done.
  In the latter case, we apply the induction hypothesis on the two premises, yielding $\dVdash{2^{(n-1)k}}{k-1}{k-1}{q} \Gamma \Rightarrow C,\Delta$ and $\dVdash{2^{(n-1)k}}{k-1}{k-1}{q} \Gamma,C \Rightarrow \Delta$.
  Then we apply \cref{0003} to eliminate inexpensive \rCut{}s; this gives $\dVdash{2^{k-1} \cdot 2^{(n-1)k}}{0}{k-1}{q} \Gamma,C \Rightarrow \Delta$ and $\dVdash{2^{k-1} \cdot 2^{(n-1)k}}{0}{k-1}{q} \Gamma \Rightarrow C,\Delta$.
  Finally, we apply \cref{0004} to obtain $\dVdash{2^{n \cdot k}}{k-1}{k-1}{q} \Gamma \Rightarrow \Delta$.
\end{proof}

Combining \cref{0003,0005} yields \cref{0006}.

\begin{corollary}\label{0006}
  If $\dVdash{n}{\ell}{k}{q+1} \Gamma \Rightarrow \Delta$, then $\dVdash{h}{k-1}{k-1}{q} \Gamma \Rightarrow \Delta$ where $h = 2^{2^{\ell} \cdot n \cdot k}$.
\end{corollary}

\begin{lemma}\label{0007}
  If $\gVdash{n}{q+1} \Gamma \Rightarrow \Delta$, then $\dVdash{h}{n-1}{n-1}{q} \Gamma \Rightarrow \Delta$ where $h = 2^n_3$.
\end{lemma}
\begin{proof}
  In a gentle proof, the rank of every cut formula is bounded by the height $n$; hence we can instantiate \cref{0006} with $\ell = k = n$, resulting in $\dVdash{h}{n-1}{n-1}{q} \Gamma \Rightarrow \Delta$ where $h = 2^{2^{n} \cdot n^2}$.
  It remains to show that $2^{2^n \cdot n^2} \leq 2^n_3$.
  To this end, it suffices to show $2^n \cdot n^2 \leq 2^{2^n}$.
  We manually verify that it is the case for $n = 0,1$; then we do an induction on $n$ to show that it holds for all $n \ge 2$.
  In the induction step, we show the following chain of inequalities:
  \[
    2^{k+1} \cdot (k+1)^2 \leq 2^{k+1} \cdot (2k)^2 \leq 8 \cdot 2^k \cdot k^2 \leq 8 \cdot 2^{2^k} \leq (2^{2^k})^2 = 2^{2^{k+1}}
  \]
  This is evident by the induction hypothesis and the fact that $k \geq 2$.
\end{proof}

\section*{Problem 2}\label{0008}
\begin{proof}[Solution]
  Suppose that $\gVdash{n}{q} \Gamma \Rightarrow \Delta$.
  Applying \cref{0007} recursively on $q$ yields $\dVdash{h}{-}{-}{0} \Gamma \Rightarrow \Delta$ where $h = 2^n_{3q}$; the omitted bounds are not important.
  Since $\qRank{A} \geq 0$ for any formula $A$, we can conclude that the resulting derivation has no cut formulas, \ie $\vdash^h_0 \Gamma \Rightarrow \Delta$.
\end{proof}

\section*{Problem 3}\label{000A}
\begin{proof}[Solution]
  Eliminating quantified cut formulas is expensive because we do not have inversion at our disposal; we have to appeal to the induction hypothesis, \ie make an expensive recursive function call, to handle implicit contraction correctly.
  
  In the intuitionistic case, negation and disjunction lack inversion in addition to quantifiers, so we modify $q$-rank to additionally count negation and disjunction, \ie $\qRank{\lnot A} = \qRank{A} + 1$ and $\qRank{A \vee B} = \max(\qRank{A},\qRank{B}) + 1$.
  Then we can prove the same statement in Problem 2.
\end{proof}

% \bibliography{bib}
% \bibliographystyle{alpha}

\end{document}
